Efficient implementation of kernel filtering

### 2007-06-22 article/Article.tex
Extended article
darcs-hash:20070622010226-fafb9-8257848d45b237122aee00f0665850732a4de374.gz

\subsection{Arbitary structuring elements}
\label{sect:MMmovingHist}
The method described in \cite{Vandroogenbroeck96.3} relies on the
simple concept of an updatable histogram. A histogram is computed for
a kernel located at the first voxel. The histogram at the neighboring
voxel can then be computed by including newly included voxels and
removing newly excluded voxels. The list of included and excluded
voxels corresponding to movement in any direction can be computed when
the structuring element is created, and the direction with the
smallest number of changes should be selected as the direction for
sweeping the kernel across the image. The erosion or dilation at each
location is computed by selecting the minimum or maximum from the
histogram. This approach is very efficient when 8 or 16 bit pixels are
used because the histogram can be represented as an array, with place
holders tracking the current maximum or minimum increasing
performance. More sophisticated histogram representations are
necessary for larger pixel types. Our implementation uses c++ maps.

@@

\subsection{Decomposition of structuring elements}
Morphological erosions and dilations are separable -- successive
dilations by orthogonal lines is equivalent to dilation by a rectangle
with sides equal to the line lengths. This means that any
hyper-rectangular structuring element can be constructed using lines
parallel to the image axes.

@@

There has been a method published recently that offers improved
performance \cite{Vandroogenbroeck2005Morphological} and the ability
to perform a line opening directly (rather than using an
erosion/dilation cascade). This method has been implemented in the
filters discussed later in the article, but performance issues are not
yet clear. The method employs histograms and therefore needs more
complex data structures when applied to higher precision data,
potentially any speed advantage.

\section{Rank filters}
Efficient implemetations of median and rank filters can be carried out
using exactly the same approach as discussed for morphological
operations with arbitary structuring elements. The method was
originally proposed in \cite{Huang79}. The implementation discussed
later supports arbitary kernel shapes and pixel types.

Rank filters are not separable. However the performance benefits
offered by separability make it worth pretending they are. If median
filtering is being used to provide robust noise filtering or
background estimation then a separable approximation is worth
testing. This concept was originally proposed in \cite{Narendra81}.

>>>>

\subsection{Arbitary structuring elements}
\label{sect:MMmovingHist}
The method described in \cite{Vandroogenbroeck96.3} relies on the
simple concept of an updatable histogram, and is often described as a
``moving histogram'' approach. A histogram is computed for a kernel
located at the first voxel. The histogram at the neighboring voxel can
then be computed by including newly included voxels and removing newly
excluded voxels. The list of included and excluded voxels
corresponding to movement in any direction can be computed when the
structuring element is created, and the direction with the smallest
number of changes should be selected as the direction for sweeping the
kernel across the image. The erosion or dilation at each location is
computed by selecting the minimum or maximum from the histogram. This
approach is very efficient when 8 or 16 bit pixels are used because
the histogram can be represented as an array, with place holders
tracking the current maximum or minimum increasing performance. More
sophisticated histogram representations are necessary for larger pixel
types. Our implementation uses c++ maps.

@@

\subsection{Decomposition of structuring elements}
\label{sect:MMdecomp}
Morphological erosions and dilations are separable -- successive
dilations by orthogonal lines is equivalent to dilation by a rectangle
with sides equal to the line lengths. This means that any
hyper-rectangular structuring element can be constructed using lines
parallel to the image axes.

@@

A method published recently that offers improved
performance \cite{Vandroogenbroeck2005Morphological} and the ability
to perform a line opening directly (rather than using an
erosion/dilation cascade). This method has been implemented in the
filters discussed later in the article, but performance issues are not
yet clear. The method employs histograms and therefore needs more
complex data structures when applied to higher precision data,
potentially any speed advantage.

\section{Rank filters}
Efficient implemetations of median and rank filters can be carried out
using exactly the same approach as discussed for morphological
operations with arbitary structuring elements. The method was
originally proposed in \cite{Huang79}. The implementation discussed
later supports arbitary kernel shapes and pixel types as well as any
choice of rank.

Rank filters are not separable. However the performance benefits
offered by separability make it worth pretending they are. If median
filtering is being used to provide robust noise filtering or
background estimation then a separable approximation is worth
testing. This concept was originally proposed in \cite{Narendra81}. In
fact it is also posible to apply the decomposition discussed in
Section \ref{sect:MMdecomp} to median filters to achieve a closer
approximation to circular kernels.

\section{Mean and variance filters}
A simplified version of the moving histogram approach can also be used
to efficiently compute the mean or variance of an arbitrarily shaped
kernel. The histogram used for morphology and rank filters is a data
structure that, once established for one pixel, can be easily updated
to represent the kernel for a neighboring pixel. The data structure
required to perform the equivalent function for mean and variance
operations is much simpler -- all that is necessary is the sum of
pixel values under the kernel (or sum of squared values for variance
calculations), and the kernel size. The sum can be updated by adding
the new values and subtracting the old and the output computed by
dividing by the kernel size.

A more efficient strategy was proposed in \cite{Crow84} for rectangular
kernels. This approach first computes an accumulator image in which
each $i,j$ is repaced by .. This accumulator image may be computed
recursively using local neighborhood values. The mean of rectangles of
any size may then be computed using accumulated values at the
rectangle corners. Figures \ref{fig:accum} amd \ref{fig:accumMean}
illustrate the 2D case and the update formulas are provided in
Equations \ref{eq:accum} and \ref{eq:accumMean}.

\begin{eqnarray}
\label{eq:accum}
b_{i,j} & = & \sum_{x \leq i, y \leq j} a_{x,y} \\ \nonumber
	& = & b_{i,j-1} + b_{i-1, j} + a_{i,j} - b_{i-1, j-1}
\end{eqnarray}
where $b_{i,j}$ is the accumulator value at location $i,j$ and $a_{i,j}$ is the input image intensity at $i,j$.

\begin{equation}
\label{eq:accumMean}
m_{i,j} = \frac{1}{wh} (b_{i+w/2,j+h/2} + b_{i-w/2,j-h/2} - b_{i-w/2,j+h/2} + b_{i+w/2,j-h/2})
\end{equation}
where $m_{i,j}$ is the output of the mean filter at location $i,j$.

\begin{figure}[htbp]
\centering
\includegraphics[scale=0.5]{accum2D}
\caption{Recursive computation of the accumulator image - value at $(i,j)$ computed using Equation \ref{eq:accum}.\label{fig:accum}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics[scale=0.5]{accumMean2d}
\caption{Computation of mean using the accumulated image - lue at $(i,j)$ computed using Equation \ref{eq:accumMean}.\label{fig:accumMean}}
\end{figure}

This approach to convolution has been used recently in face detection
applications \cite{Viola2004}, where rectangular filters of many
different sizes and shapes were needed.

\section{ITK implementation}
